Fisher vector places: learning compact descriptors for place recognition

### 2014-11-14 cvpr15review.tex
final commit before submission

% usepackages by Petr
\usepackage{epstopdf}
\usepackage{enumerate}
\usepackage{subfigure}
\usepackage{caption}
\usepackage{tabularx}
\usepackage[table]{xcolor}
\usepackage{hyphenat}

% Include other packages here, before hyperref.

@@

%%%%%%%%% ABSTRACT
\begin{abstract}
   \noindent
   The goal of this work is to localize a query photograph by finding other images depicting the same place in a large geotagged image database. The contributions of this work are three fold. First, we learn a {\em discriminative} yet {\em compact} descriptor of each image in the database. This is achieved by applying exemplar support vector machine (e-SVM) learning to compact Fisher vector descriptors extracted from database images. Secondly, we analyze the exemplar support vector machine objective and show that the learnt hyperplane can be interpreted as a new descriptor that replaces the original positive example and is re-weighted to increase its separation from the negative data. Finally, based on this analysis, we demonstrate that the learnt and re-normalized descriptor could be directly used for matching, thus avoiding the need for expensive and tedious calibration typically needed for exemplar support vector machine methods. Place recognition results are shown on two image datasets of Google street-view images from Pittsburgh  and demonstrate the learnt representation consistently improves over the standard Fisher vector descriptors at different target dimensions. 
\end{abstract}

%%%%%%%%% BODY TEXT
\section{Introduction}
\noindent
The goal of this work is to localize a query image by matching to a large database of geotagged street-level imagery.
This is an important problem with practical applications in robotics, augmented reality or navigation. This task is however very difficult. It is hard to distinguish different places, e.g.\ streets in a city, from each other. The imaged appearance of a place can change drastically due to factors such as  viewpoint, illumination or even changes over time.
%
Finally, with the emergence of planet-scale geotagged image collections, such as Google Street-view, the image databases are becoming very large. We estimate a single country like France is covered by more than 60 million street-level panoramas. Hence the fundamental challenge in place recognition lies now in designing robust, discriminative, yet compact, image representations.

In this work we build on the method of Gronat {\it et al.}~\cite{Gronat13} who represent each image in the database by a per-location classifier that is trained to discriminate each place from other places in the database. At query time, the query image is classified by all per-location classifiers and assigned to a place with the highest classification score. The training of each classifier is performed using the per-exemplar support vector machine (e-SVM)~\cite{Malisiewicz11}, which takes the positive image as a single positive example and other far away images in the database as negative examples. 
The exemplar SVM is well suited for this task as street-level image collections typically contain only one or at most a hand-full of images depicting the same place. The intuition is that the exemplar SVM can learn the important features that distinguish the particular place from other similar places  in the database.  While the results of~\cite{Gronat13} are promising they suffer from two important drawbacks. First, the learnt place specific representation is not compact, which prohibits its application to planet-scale street-level collections that are now becoming available~\cite{Klinger13}. Second, the per-exemplar classifiers require careful and time-consuming calibration.

In this work we address both these issues. 
First, we apply the exemplar SVM training to compact Fisher vector~\cite{Jegou2011,Perronnin2010} image descriptors, which results in a {\em discriminative} yet {\em compact} representation of each image in the database. Second, to avoid the expensive classifier calibration, we analyze the exemplar SVM cost and show that the learnt hyperplane can be interpreted as a new descriptor that replaces the original positive example and is re-weighted to increase its separation from the negative data. As a result of this analysis, we demonstrate that after an appropriate normalization of the new re-weighted descriptor no further calibration is necessary. We show improved results on place recognition using learnt compact Fisher vector descriptors~\cite{Jegou2011} of different dimensionality.
%-------------------------------------------------------------------------
\section{Related work} 
\label{sec:related}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Fisher vector image representations.} %\vspace{-3mm}
    Fisher vector image representations have recently demonstrated excellent performance for a number of visual recognition tasks~\cite{Chatfield11,Jegou12,Krapac2011,Simonyan2013}. They are specially suited for retrieval applications since they are robust to image appearance variations and capture richer image statistics than the simple bag-of-visual-words (BOW) aggregation. However, the raw extracted Fisher vectors are typically high-dimensional, e.g.\ with 32,768 non-sparse dimensions, which is impractical for large-scale visual recognition and indexing applications. Hence, their dimensionality is often reduced by principal component analysis (PCA) and further quantized for efficient indexing using, e.g.\ a product quantizer~\cite{Jegou12}.  Other recent work has demonstrated improved performance in a face recognition application by finding discriminative projection using a large amount of training face data~\cite{Simonyan2013}. Our work is complementary to these methods as it operates on the projected low-dimensional descriptor and further learns discriminative re-weighting of the descriptor specific to each image in the database using per-exemplar support vector machine~\cite{Malisiewicz11}.

@@

\paragraph{Contributions.} %\vspace{-3mm}
    The contributions of this work are threefold.
    First, we analyze the exemplar support vector machine objective and show that the learnt hyperplane can be interpreted as a new descriptor that replaces the original positive example and is re-weighted to increase its separation from the negative data. Secondly, we demonstrate that after an appropriate normalization of the new re-weighted descriptor no further calibration is necessary. Finally, we apply e-SVM training to compact Fisher vector descriptors for large-scale place recognition resulting in a {\em discriminative} yet {\em compact} representation of each image in the database. Place recognition results are shown on a dataset of 25k images of Pittsburgh and demonstrate the learnt representation consistently improves over the standard Fisher vector descriptors at different target dimensions.

\section{Learning compact place descriptors using per-exemplar SVM}
\label{sec:perExemplar}
   Each database image $j$ is represented by its L2-normalized Fisher vector $\Phi_j$. The goal is to learn a set of new L2-normalized Fisher vectors $\Psi_j$, one per each database image.
   % such that at query time, 
   At query time, given the Fisher vector $\Phi_q$ of an unknown query image, we retrieve the database image depicting the same location by finding the image $j^*$ with the highest score measured by a dot product
   %
   \begin{equation}
       j^*=\operatorname*{arg\;max}_{j} \Phi_q^T \Psi_j. 
       \label{eq:class}
   \end{equation}
   %  
   In other words, the aim is to replace each original database Fisher vector $\Phi_j$ with a new vector $\Psi_j$ that is more discriminative in the sense of separation from descriptors of images depicting other places. Inspired by~\cite{Gronat13}, we investigate applying the exemplar support vector machine (e-SVM)~\cite{Malisiewicz11} for this task. e-SVM learns a linear classifier $w_j^T\Phi+b_j$ given the descriptor $\Phi_j^+$ of place $j$ as a single positive example (with target label +1) and a large number of negative descriptors $\mathcal N_j$ from other places in the database (with target labels -1). The intuition of the exemplar SVM training~\cite{Malisiewicz11} is that the learnt weight vector $w_j$ will give a higher weight to the dimensions of the descriptor that are discriminative and will down-weight dimensions that are non-discriminative with respect to the negative training data collected from other far-away places.  The optimal $w_j$ and $b_j$ are obtained by minimizing the following objective  
      %
      \begin{align}
         \label{eq:obj}
         ||w_j||^{2} &+C_1 \cdot h
            \left(
               w_j^T\Phi^+_j+b_j
            \right)  \\ \nonumber
            &+C_2\sum_{\Phi\in \mathcal N_j}h
            \left(
               -w_j^T\Phi-b_j
            \right),    
      \end{align}
      %
   where $\Phi^+_j$ is the descriptor of the place $j$ as the positive data point, $\Phi$ are Fisher descriptors from negative training data $\mathcal N_j$ and $h$ is the hinge loss, $h(y) = \max(1-y,0)$. Note that the first term in~(\ref{eq:obj}) is the regularizer, the second term is the loss on the positive data weighted by scalar parameter $C_1$ and the third term is the loss on the negative data weighted by scalar parameter $C_2$. The objective is convex and can be minimized with respect to $w_j$ and $b_j$ using standard software packages such as~\cite{libsvm}. A key advantage is that the per-exemplar classifier for each place can be trained independently and hence the learning can be heavily parallelized. The downside of the independent training for each positive example is that the resulting scores have to be calibrated with respect to each other on additional data~\cite{Gronat13,Malisiewicz11}.

   \subsection{Analysis of per-exemplar SVM objective} %\vspace{-0.2cm} 
      In this section, we analyze the exemplar SVM objective~(\ref{eq:obj}) and show the learnt and {\em re-normalized} weight vector $w_j$ can be interpreted as a new descriptor $\Psi_j$ that replaces the original positive training descriptor $\Phi^+_j$.  In particular, we show first that when the weight $C_2$ of the negative data in objective~(\ref{eq:obj}) goes to zero and the learnt $\Psi_j$ is identical to the original positive training data point $\Phi_j^+$.  Second, when $C_2>0$, the learnt $\Psi_j$ moves away from the positive $\Phi_j^+$ to increase its separation from the negative data.
      Details are given next.
            %
            \begin{figure}[t]
               \begin{center}
                  \subfigure[$C_2>0$ ]{
                     \centering
                     \includegraphics[width=0.45\linewidth]{imgs/figSVM01tight}
                  }
                  %        $C_2 \rightarrow 0$
                  \subfigure[$C_2 \rightarrow 0$]{
                     \centering
                     \includegraphics[width=0.45\linewidth]{imgs/figSVM02tight}
                  }
                  \caption{
                     {\bf An illustration of the effect of decreasing parameter $C_2$ in the exemplar support vector machine objective.} 
                     The positive exemplar $\Phi_j^+$ is shown in green. The negative data points are shown in red. All training data is L2 normalized to lie on a hyper-sphere. (a) For $C_2>0$, the normal $w_j$ of the optimal hyper-plane moves away from the direction given by the positive example $\Phi^+$ in a manner that reduces the loss on the negative data.  (b) As the parameter $C_2$ decreases the learnt $w_j$ becomes parallel to the positive training example $\Phi_j^+$ and its magnitude $||w_j||$ goes to 0.
                  }
                  \label{fig:C2effect}
               \end{center}
            \end{figure}
            %
      \paragraph{Case I: $C_2\rightarrow 0$.}
         The goal is to show that when the weight  $C_2$ of the negative data in objective~\eqref{eq:obj} goes towards zero, the resulting hyperplane vector $w_j$ is parallel with the vector of positive training descriptor $\Phi_j^+$. When $w_j$ is normalized to have unit L2 norm the two vectors are identical. First, let us decompose $w$ into parallel and orthogonal part with respect to the positive training data point $\Phi^+$ (in the following we omit index $j$ for brevity), i.e. $w=w^{\perp}+w^{||}$, where $(w^{\perp})^T \Phi^+ = 0$. Next, we observe that when the weight of the negative data diminishes ($C_2\rightarrow 0$), any non-zero component $w^{\perp}$ will increase the value of the objective. As a result, for $C_2\rightarrow 0$ the objective is minimized by $w^{||}$, i.e.\ the optimal $w$ is parallel with $\Phi^+$.

@@

      \paragraph{Case II: $C_2>0$.}
         When the weight $C_2$ of the negative data in the objective~\eqref{eq:decomposed} increases the direction of the optimal $w$ will be different from $w^{||}$ and will change to take into account the loss on the negative data points. Explicitly writing the hinge-loss $h(x) = \max(1-x,0)$ in the last term of~\eqref{eq:decomposed}, we see that $w$ will move in the direction that reduces $\sum_{\Phi\in \mathcal N}\max \left(1+w^T \Phi + b ,0 \right)$, i.e.\ that reduces the dot product $w^T \Phi$ on the negative examples that are active (support vectors).

   \subsection{Interpreting normalized $w$ as a new descriptor} %\vspace{-0.2cm} 
      The above analysis demonstrates that as $C_2$ decreases the normal of the optimal hyperplane $w$ that separates the positive exemplar $\Phi^+$ from negative data becomes parallel with $\Phi^+$, as shown in figure \ref{fig:C2effect}. As $C_2$ increases, the normal $w$ of the optimal hyper-plane moves away from the direction given by the positive example $\Phi^+$ in a manner that reduces the loss on the negative data. 
      This suggests that the learnt $w$ could be interpreted as a modified positive example $\Phi^+$, re-weighted to emphasize directions that separate $\Phi^+$ from the negative data. As discussed above $w$ is not normalized. As we wish to measure the similarity between descriptors by (the cosine of) their angle given by equation \eqref{eq:class}, additional normalization of the learnt $w$ is necessary. Hence we define the new descriptor $\Psi_j$ as the normalized hyperplane normal $w_j$
         %
         \begin{equation}
            \label{eq:normalization}
            \Psi_j=\dfrac{w_j}{||w_j||}.
         \end{equation}
      \textcolor{petr}{%TODO: \emph{we should't speak about calibration, we shoud address not ignoring $b$}\\
%      Measuring the similarity between descriptors by their normalized cross-correlation 
         Note again that for L2 normalized query descriptor $\Phi_q$, evaluating $\Phi_q^T\Psi_j$ in eq.~\eqref{eq:class} corresponds to computing the~normalized cross-correlation between $\Phi_q$ and $\Psi_j$. This was found to work well in retrieval (see e.g.~\cite{Sivic03}) or matching whitened HOG descriptors \cite{Gharbi12, Hariharan12}, but has not yet been used in conjunction with SVMs.    
         Note that while the bias term $b$ is not directly used in the construction of $\Psi_j$, the bias affects the learnt optimal $w_j$ as it appears in the SVM objective~\eqref{eq:obj}. 
         % when learning the optimal SVM hyper-plane $w_j$.           
%         even the new descriptor $\Psi_j$ is obtained by normalization of the learnt hyperplane $w_j$, the bias $b$ form equation~\eqref{eq:decomposed} is not ignored. Indeed, the bias term in convex objective~\eqref{eq:obj} is affecting the learnt the hyperplane.
%         %\\
         %The proposed re-normalization procedure can be interpreted as a simple form of affine calibration where the uncalibrated SVM score $s(x)=w^T x+b$ is transformed into a calibrated score $f(x)=\alpha s(x)+\beta$, where $\alpha = 1/||w||$ and $\beta = -b\alpha$. Note that the bias term b is not ignored but used to compute $f(x)$. The intuition is that for L2 normalized descriptors x computing the calibrated score $f(x)$ corresponds to computing the normalized cross-correlation between vectors w and x. This was found to work well in retrieval (see e.g. Sivic and Zisserman, ICCV 2003 [32]) or matching whitened HOG descriptors [11,13] (see e.g. [Doersch et al., Mid-Level Visual Element Discovery as discriminative Mode Seeking, NIPS 2013]), but has not yet been used in conjunction with SVMs as done in this work. Note also that the affine calibration can be computed offline and does not need any additional computation or storage at query time compared to the p-value calibration of Gronat et al. [12].
      }

@@

   \subsection{Experimental set-up}
      We perform experiments on a database of Google Street View images of Pittsburgh downloaded from the Internet. The data contains panoramas covering roughly an area of $1.3 \times 1.2 \; {\rm km}^2$.       
      Similar to~\cite{Chen11}, for each panorama we generate 12 overlapping perspective views corresponding to two different elevation angles to capture both the street-level scene and the building fa\c{c}ades, resulting in a total of 24 perspective views each with $90^\circ$ FOV and resolution of $960 \times 720$ pixels.
      For evaluation we have used two versions of this data. The first one was obtained from the authors of~\cite{Gronat13} ($25$k images). In the second version we download additional images to increase the dataset size to $55$k images. 
      As a query set with known ground truth GPS positions, we use the 8999 panoramas from the Google Street View research dataset. This dataset covers approximately the same area, but has been captured at a different time, and depicts the same places from different viewpoints and under different illumination conditions. For each test panorama, we generate a set of perspective images as described above. Finally, we randomly select out of all generated perspective views a subset of 4k images, which is used as a test set to evaluate the performance of the proposed approach.
      Since all the query images have associated GPS location we can compute their spatial distance from the database images returned by the matching method. We consider a query image to be correctly localized if the retrieved database image lies within a perimeter of $20m$ from the location of the query.
      %
      \begin{figure}[t!]
         \centering
         \includegraphics[width=\linewidth]{imgs/plotPitt25kSURF}    
         \caption{
            \textbf{Evaluation on Pittsburgh 25k \cite{Gronat13} dataset.} The fraction of correctly recognized queries (recall@K, y-axis) vs. the number of top $K$ retrieved database images for different Fisher vector dimensions. The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions.
         }
         \label{fig:recall}
      \end{figure}

@@

   \subsection{Results} 
      \textcolor{petr}{
         The experiments are performed on two datasets (Pittsburgh 25k and Pittsburgh 55k) on several target Fisher vector dimensions $d\in\{128,512,2048\}$. The proposed approach can be also applied to other descriptors beyond Fisher vectors. To demonstrate this we have applied the proposed method to the bag-of-visual-words (BOW).
         \\
         For each method we measure performance using the percentage of correctly recognized queries (Recall) similarly to, e.g.,~\cite{Chen11,Knopp2010,Sattler-BMVC12}. The query is correctly localized if at least one of the top $K$ retrieved database images is within $20$ meters from the ground truth position of the query. Results are shown for different values of $K$ in tables~\ref{tab:recallBOW},~\ref{tab:recall} and \ref{tab:recall2}. For the Pittsburgh 25k we also show results in the form of a curve in figure~\ref{fig:recall}. The results clearly demonstrate the benefits of the learnt descriptors with respect to the standard Fisher vectors for all target dimensions and lengths of shortlist $K$. The benefits of discriminative learning are specially prominent for low-dimensional compact descriptors ($d=128$). Figure~\ref{fig:images} shows examples of place recognition results.
      }

      % For both databases (Pittsburgh 25k and Pittsburgh 55k) we compare results of our method (FV e-SVM) to two baselines: standard bag-of-visual-words baseline (BOW) and raw Fisher vector matching without learning (FV).

      % We perform experiments on several target Fisher vector dimensions $d\in\{128,512,2048\}$. For each method we measure performance using the percentage of correctly recognized queries (Recall) similarly to, e.g.,~\cite{Chen11,Knopp2010,Sattler-BMVC12}. The query is correctly localized if at least one of the top $K$ retrieved database images is within $20$ meters from the ground truth position of the query. Results are shown for different values of $K$ in tables~\ref{tab:recall} and \ref{tab:recall2}. For the Pittsburgh 25k we also show results in the form of a curve in figure~\ref{fig:recall}. The results clearly demonstrate the benefits of the learnt descriptors with respect to the standard Fisher vectors for all target dimensions and lengths of shortlist $K$. The benefits of discriminative learning are specially prominent for low-dimensional compact descriptors ($d=128$). 
      % The proposed method also significantly outperforms the bag-of-visual-words baseline. Figure~\ref{fig:images} shows examples of place recognition results.

>>>>

% usepackages by Petr
\usepackage{epstopdf}
\usepackage{enumerate}
\usepackage{subfigure}
\usepackage{caption}
\usepackage{tabularx}
\usepackage[table]{xcolor}
\usepackage{hyphenat}
\renewcommand{\baselinestretch}{0.986}
% Include other packages here, before hyperref.

@@

%%%%%%%%% ABSTRACT
\begin{abstract}
   \noindent
   The goal of this work is to localize a query photograph by finding other images depicting the same place in a large geotagged image database. The contributions of this work are three fold. First, we learn a {\em discriminative}, yet {\em compact}, descriptor of each image in the database. This is achieved by applying exemplar support vector machine (e-SVM) learning to compact Fisher vector descriptors extracted from database images. Secondly, we analyze the exemplar support vector machine objective and show that the learnt hyperplane can be interpreted as a new descriptor that replaces the original positive example and is re-weighted to increase its separation from the negative data. Finally, based on this analysis, we demonstrate that the learnt and re-normalized descriptor can be directly used for matching, thus avoiding the need for expensive and tedious calibration typically needed for exemplar support vector machine methods. Place recognition results are shown on two image datasets of Google street-view images from Pittsburgh  and demonstrate the learnt representation consistently improves over the standard Fisher vector descriptors at different target dimensions. 
\end{abstract}

%%%%%%%%% BODY TEXT
\section{Introduction}
   \noindent
   The goal of this work is to localize a query image by matching to a large database of geotagged street-level imagery.
   This is an important problem with practical applications in robotics, augmented reality or navigation. This task is however very difficult. It is hard to distinguish different places, e.g.\ streets in a city, from each other. The image appearance of a place can change drastically due to factors such as  viewpoint, illumination or even changes over time.
   %
   Finally, with the emergence of planet-scale geotagged image collections, such as Google Street-view, the image databases are becoming very large. We estimate a single country like France is covered by more than 60 million street-level panoramas. Hence the fundamental challenge in place recognition lies now in designing robust, discriminative, yet compact, image representations.

   In this work we build on the method of Gronat {\it et al.}~\cite{Gronat13} who represent each image in the database by a per-location classifier that is trained to discriminate each place from other places in the database. At query time, the query image is classified by all per-location classifiers and assigned to a place with the highest classification score. The training of each classifier is performed using the per-exemplar support vector machine (e-SVM)~\cite{Malisiewicz11}, which takes the positive image as a single positive example and other far away images in the database as negative examples. 
   The exemplar SVM is well suited for this task as street-level image collections typically contain only one or at most a hand-full of images depicting the same place. The intuition is that the exemplar SVM can learn the important features that distinguish the particular place from other similar places  in the database.  While the results of~\cite{Gronat13} are promising they suffer from two important drawbacks. First, the learnt place specific representation is not compact, which prohibits its application to planet-scale street-level collections that are now becoming available~\cite{Klinger13}. Secondly, the per-exemplar classifiers require careful and time-consuming calibration.

   In this work we address both these issues. 
   First, we apply the exemplar SVM training to compact Fisher vector~\cite{Jegou2011,Perronnin2010} image descriptors, which results in a {\em discriminative}, yet {\em compact}, representation of each image in the database. Secondly, to avoid the expensive classifier calibration, we analyze the exemplar SVM cost and show that the learnt hyperplane can be interpreted as a new descriptor that replaces the original positive example and is re-weighted to increase its separation from the negative data. As a result of this analysis, we demonstrate that after an appropriate normalization of the new re-weighted descriptor no further calibration is necessary. We show improved results on place recognition using learnt compact Fisher vector descriptors~\cite{Jegou2011} of different dimensionality.

\section{Related work} 
   \label{sec:related}

@@

   \paragraph{Fisher vector image representations.} %\vspace{-3mm}
      Fisher vector image representations have recently demonstrated excellent performance for a number of visual recognition tasks~\cite{Chatfield11,Jegou12,Krapac2011,Simonyan2013}. They are specially suited for retrieval applications since they are robust to image appearance variations and capture richer image statistics than the simple bag-of-visual-words (BOW) aggregation. However, the raw extracted Fisher vectors are typically high-dimensional, e.g.\ with 32,768 non-sparse dimensions, which is impractical for large-scale visual recognition and indexing applications. Hence, their dimensionality is often reduced by principal component analysis (PCA) and further quantized for efficient indexing using e.g.\ a product quantizer~\cite{Jegou12}.  Other recent work has demonstrated improved performance in a face recognition application by finding discriminative projection using a large amount of training face data~\cite{Simonyan2013}. Our work is complementary to these methods as it operates on the projected low-dimensional descriptor and further learns discriminative re-weighting of the descriptor specific to each image in the database using per-exemplar support vector machine~\cite{Malisiewicz11}.

@@

   \paragraph{Contributions.} %\vspace{-3mm}
      The contributions of this work are threefold.
      First, we analyze the exemplar support vector machine objective and show that the learnt hyperplane can be interpreted as a new descriptor that replaces the original positive example and is re-weighted to increase its separation from the negative data. Secondly, we demonstrate that after an appropriate normalization of the new re-weighted descriptor no further calibration is necessary. Finally, we apply e-SVM training to compact Fisher vector descriptors for large-scale place recognition resulting in a {\em discriminative}, yet {\em compact}, representation of each image in the database. Place recognition results are shown on a dataset of 25k images of Pittsburgh and demonstrate the learnt representation consistently improves over the standard Fisher vector descriptors at different target dimensions.

\section{Learning compact place descriptors using per-exemplar SVM}
   \label{sec:perExemplar}
   Each database image $j$ is represented by its L2-normalized Fisher vector $\Phi_j$. The goal is to learn a set of new L2-normalized Fisher vectors $\Psi_j$, one per each database image.
   % such that at query time, 
   At query time, given the Fisher vector $\Phi_q$ of an unknown query image, we retrieve the database image depicting the same location by finding the image $j^*$ with the highest score measured by a dot product
   %
   \begin{equation}
       j^*=\operatorname*{arg\;max}_{j} \Phi_q^T \Psi_j. 
       \label{eq:class}
   \end{equation}
   %  
   In other words, the aim is to replace each original database Fisher vector $\Phi_j$ with a new vector $\Psi_j$ that is more discriminative in the sense of separation from descriptors of images depicting other places. Inspired by~\cite{Gronat13}, we investigate applying the exemplar support vector machine (e-SVM)~\cite{Malisiewicz11} for this task. e-SVM learns a linear classifier $w_j^T\Phi+b_j$ given the descriptor $\Phi_j^+$ of place $j$ as a single positive training example (with target label +1) and a large number of negative descriptors $\mathcal N_j$ from other places in the database (with target labels -1). The intuition of the exemplar SVM training~\cite{Malisiewicz11} is that the learnt weight vector $w_j$ will give a higher weight to the dimensions of the descriptor that are discriminative and will down-weight dimensions that are non-discriminative with respect to the negative training data collected from other far-away places.  The optimal $w_j$ and $b_j$ are obtained by minimizing the following objective  
   %
   \begin{align}
      \label{eq:obj}
      ||w_j||^{2} &+C_1 \cdot h
         \left(
            w_j^T\Phi^+_j+b_j
         \right)  \\ \nonumber
         &+C_2\sum_{\Phi\in \mathcal N_j}h
         \left(
            -w_j^T\Phi-b_j
         \right),    
   \end{align}
   %
   where $\Phi^+_j$ is the descriptor of the place $j$ as the positive data point, $\Phi$ are Fisher descriptors from negative training data $\mathcal N_j$ and $h$ is the hinge loss, $h(y) = \max(1-y,0)$. Note that the first term in~(\ref{eq:obj}) is the regularizer, the second term is the loss on the positive data weighted by scalar parameter $C_1$ and the third term is the loss on the negative data weighted by scalar parameter $C_2$. The objective is convex and can be minimized with respect to $w_j$ and $b_j$ using standard software packages such as~\cite{libsvm}. A key advantage is that the per-exemplar classifier for each place can be trained independently and hence the learning can be heavily parallelized. The downside of the independent training for each positive example is that the resulting scores have to be calibrated with respect to each other on additional data~\cite{Gronat13,Malisiewicz11}.

   \subsection{Analysis of per-exemplar SVM objective} %\vspace{-0.2cm} 
      In this section, we analyze the exemplar SVM objective~(\ref{eq:obj}) and show the learnt and {\em re-normalized} weight vector $w_j$ can be interpreted as a new descriptor $\Psi_j$ that replaces the original positive training descriptor $\Phi^+_j$.  In particular, we show first that when the weight $C_2$ of the negative data in objective~(\ref{eq:obj}) goes to zero the learnt $\Psi_j$ is identical to the original positive training data point $\Phi_j^+$.  Second, when $C_2>0$, the learnt vector $\Psi_j$ moves away from the positive vector  $\Phi_j^+$ to increase its separation from the negative data.
      Details are given next.
      %
      \begin{figure}[t]
         \begin{center}
            \subfigure[$C_2>0$ ]{
               \centering
               \includegraphics[trim = 4mm 0mm 7mm 3mm, clip=true, width=0.47\linewidth]{imgs/figSVM01tight}
            }
            %        $C_2 \rightarrow 0$
            \subfigure[$C_2 \rightarrow 0$]{
               \centering
               \includegraphics[trim = 3mm 0mm 8mm 3mm, clip=true, width=0.47\linewidth]{imgs/figSVM02tight}
            }
            \caption{
               {\bf An illustration of the effect of decreasing parameter $C_2$ in the exemplar support vector machine objective.} 
               The positive exemplar $\Phi_j^+$ is shown in green. The negative data points are shown in red. All training data is L2 normalized to lie on a hyper-sphere. (a) For $C_2>0$, the normal $w_j$ of the optimal hyper-plane moves away from the direction given by the positive example $\Phi^+$ in a manner that reduces the loss on the negative data.  (b) As the parameter $C_2$ decreases the learnt $w_j$ becomes parallel to the positive training example $\Phi_j^+$ and its magnitude $||w_j||$ goes to 0.
            }
            \label{fig:C2effect}
         \end{center}
      \end{figure}
      %
      \paragraph{Case I: $C_2\rightarrow 0$.}
         The goal is to show that when the weight  $C_2$ of the negative data in objective~\eqref{eq:obj} goes towards zero, the resulting hyperplane vector $w_j$ is parallel with the vector of positive training descriptor $\Phi_j^+$. When $w_j$ is normalized to have unit L2 norm the two vectors are identical. First, let us decompose $w$ into parallel and orthogonal part with respect to the positive training data point $\Phi^+$ (in the following we omit index $j$ for brevity), i.e. $w=w^{\perp}+w^{||}$, where $(w^{\perp})^T \Phi^+ = 0$. Next, we observe that when the weight of the negative data diminishes ($C_2\rightarrow 0$), any non-zero component $w^{\perp}$ will increase the value of the objective. As a result, for $C_2\rightarrow 0$ the objective is minimized by $w^{||}$, i.e.\ the optimal $w$ is parallel with $\Phi^+$.

@@

      \paragraph{Case II: $C_2>0$.}
         When the weight $C_2$ of the negative data in the objective~\eqref{eq:decomposed} increases the direction of the optimal $w$ will be different from $w^{||}$ and will change to take into account the loss on the negative data points. Explicitly writing the hinge-loss $h(x) = \max(1-x,0)$ in the last term of~\eqref{eq:decomposed}, we see that $w$ will move in the direction that reduces $\sum_{\Phi\in \mathcal N}\max \left(1+w^T \Phi + b ,0 \right)$, i.e.\ that reduces the dot product $w^T \Phi$ on the negative examples that are active (support vectors).
         %

   \subsection{Interpreting normalized $w$ as a new descriptor} %\vspace{-0.2cm} 
      The above analysis demonstrates that as $C_2$ decreases the normal of the optimal hyperplane $w$ that separates the positive exemplar $\Phi^+$ from negative data becomes parallel with $\Phi^+$, as shown in figure \ref{fig:C2effect}. As $C_2$ increases, the normal $w$ of the optimal hyper-plane moves away from the direction given by the positive example $\Phi^+$ in a manner that reduces the loss on the negative data. 
      This suggests that the learnt $w$ could be interpreted as a modified positive example $\Phi^+$, re-weighted to emphasize directions that separate $\Phi^+$ from the negative data. As discussed above $w$ is not normalized. As we wish to measure the similarity between descriptors by (the cosine of) their angle given by equation \eqref{eq:class}, additional normalization of the learnt $w$ is necessary. Hence we define the new descriptor $\Psi_j$ as the normalized hyperplane normal~$w_j$
      %
      \begin{equation}
         \label{eq:normalization}
         \Psi_j=\dfrac{w_j}{||w_j||}.
      \end{equation}

      Note again that for L2 normalized query descriptor $\Phi_q$, evaluating $\Phi_q^T\Psi_j$ in eq.~\eqref{eq:class} corresponds to computing the~normalized cross-correlation between $\Phi_q$ and $\Psi_j$. This was found to work well in retrieval (see e.g.~\cite{Sivic03}) or matching whitened HOG descriptors \cite{Gharbi12, Hariharan12}, but has not yet been used in conjunction with SVMs.    
      Note that while the bias term $b$ is not directly used in the construction of $\Psi_j$, the bias affects the learnt optimal $w_j$ as it appears in the SVM objective~\eqref{eq:obj}.

@@

   \subsection{Experimental set-up}
      We perform experiments on a database of Google Street View images of Pittsburgh downloaded from the Internet. The data contains panoramas covering roughly an area of $1.3 \times 1.2 \; {\rm km}^2$.       
      Similar to~\cite{Chen11}, for each panorama we generate 12 overlapping perspective views corresponding to two different elevation angles to capture both the street-level scene and the building fa\c{c}ades, resulting in a total of 24 perspective views each with $90^\circ$ FOV and resolution of $960 \times 720$ pixels.
      For evaluation we have used two versions of this data. The first one was obtained from the authors of~\cite{Gronat13} ($25$k images). In the second version we download additional images to increase the dataset size to $55$k images. 
      As a query set with known ground truth GPS positions, we use the 8999 panoramas from the Google Street View research dataset. This dataset covers approximately the same area, but has been captured at a different time, and depicts the same places from different viewpoints and under different illumination conditions. For each test panorama, we generate a set of perspective images as described above. Finally, we randomly select out of all generated perspective views a subset of 4k images, which is used as a test set to evaluate the performance of the proposed approach.
      Since all the query images have associated GPS location we can compute their spatial distance from the database images returned by the matching method. We consider a query image to be correctly localized if the retrieved database image lies within a perimeter of $20m$ from the location of the query.
      %
      \begin{figure}[t!]
         \centering
         \includegraphics[trim = 7mm 0mm 12mm 6mm, clip=true, width=1.0\linewidth]{imgs/plotPitt25kSURF}    
         \caption{
            \textbf{Evaluation on Pittsburgh 25k \cite{Gronat13} dataset.} The fraction of correctly recognized queries (recall@K, y-axis) vs. the number of top $K$ retrieved database images for different Fisher vector dimensions. The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$ and all~dimensions.
         }
         \label{fig:recall}
      \end{figure}

@@

   \subsection{Results} 
      The experiments are performed on two datasets (Pittsburgh 25k and Pittsburgh 55k) on several target Fisher vector dimensions $d\in\{128,512,2048\}$. The proposed approach can be also applied to other descriptors beyond Fisher vectors. To demonstrate this we have applied the proposed method to the bag-of-visual-words (BOW).

      For each method we measure performance using the percentage of correctly recognized queries (Recall) similarly to, e.g.,~\cite{Chen11,Knopp2010,Sattler-BMVC12}. The query is correctly localized if at least one of the top $K$ retrieved database images is within $20$ meters from the ground truth position of the query. Results are shown for different values of $K$ in tables~\ref{tab:recallBOW},~\ref{tab:recall} and \ref{tab:recall2}. For the Pittsburgh 25k we also show results in the form of a curve in figure~\ref{fig:recall}. The results clearly demonstrate the benefits of the learnt descriptors with respect to the standard Fisher vectors for all target dimensions and lengths of shortlist $K$. The benefits of discriminative learning are specially prominent for low-dimensional compact descriptors ($d=128$). Figure~\ref{fig:images} shows examples of place recognition results.
